---
title: "ASICS"
output:
  pdf_document: default
  html_document: default
date: "2026-09-16"
---

```{r setup, include=FALSE}
library(tidyverse)
library(magrittr)
library(ASICS)
library(pheatmap)
```

## ASICS vignette example


```{r cars}
current_path <- system.file("extdata", "example_spectra", package = "ASICS")
spectra_data <- importSpectraBruker(current_path)
spectra_obj <- createSpectra(spectra_data)
to_exclude <- matrix(c(4.5, 10), ncol = 2)
resASICS <- ASICS(spectra_obj, exclusion.areas = to_exclude, verbose = FALSE)

```

## Load our spectra

we put ppm.grid as NULL, in which case it defaults to the pure_library ppm grid which goes from -0.5 to 10. It seems that all of the samples have slightly different ppm grids (same length, but different offsets), which messes with the function.
```{r pressure}
#media_spectra <- importSpectraBruker('/Users/b246357/Documents/uni/thesis/media_extraction_comparison/Lars260911', which.spectra = NULL, ppm.grid = as.numeric(colnames(signalintegral))) %>% select(-c('21', '91')) # don't load 2D spectra
#spectral_grid <- spectra$Lars260911$cs # from metabodeconplus

media_spectra <- importSpectraBruker('/Users/b246357/Documents/uni/thesis/media_extraction_comparison/only_1d_spectra', which.spectra = NULL, ppm.grid = NULL) # don't load 2D spectra

```



```{r}
spectra_norm <- normaliseSpectra(media_spectra, type.norm = "pqn")
spectra_align <- alignSpectra(spectra_norm)
spectra_obj <- createSpectra(spectra_align)
```


```{r}
to_exclude <- matrix(c(-3, -0.2, 8, 13, 4.8, 6), ncol = 2, byrow = TRUE)
ASICS_results <- ASICS(spectra_obj, exclusion.areas = to_exclude)
```


```{r}
plot(ASICS_results, idx = 1, add.metab = "Lactate")
```


```{r}
plot(ASICS_results, idx = 1, add.metab = "Pyruvic-Acid")
```



PCA
```{r}
asics_pca <- ASICS_results@quantification %>% t() %>% prcomp(scale. = F)
extraction_method <- c('E', 'K', 'E', 'E', 'K', 'K', 'K', 'E','E')
media_type = c('+cell', '+cell', '+cell', '-cell', '-cell', '-cell', '+cell', '-cell', '+cell')

annotation_matrix <- bind_cols(extraction_method = factor(extraction_method), media_type = factor(media_type)) %>% data.frame()
rownames(annotation_matrix) <- ASICS_results@sample.name
asics_pca_df <- asics_pca$x %>% as.data.frame() %>% bind_cols(annotation_matrix)
asics_pca_df %>% ggplot(aes(x=PC1, y=PC2, col=media_type, shape=extraction_method))+geom_point(size=5)

```


heatmap
```{r}
metabolite_mat <- as.matrix(ASICS_results@quantification)
metabolite_var <- apply(metabolite_mat, 1, var, na.rm = TRUE)

# Keep top 30 most variable metabolites
top <- names(sort(metabolite_var, decreasing = TRUE))[1:30]

mat_top <- metabolite_mat[top, , drop = FALSE]


pheatmap::pheatmap(
  #metabolite_mat,
  mat_top,
  annotation_col = annotation_matrix,
  cluster_cols = TRUE,
  cluster_rows = TRUE,
  border_color = NA,
  fontsize_row = 7
)
```

looks like number 7 has the highest sum but it should have the lowest?
```{r}
metabolite_mat |> apply(2, sum)
```


Try Anders' old metabolomics data
```{r}
nmr_csv <- read_csv('/Users/b246357/Documents/uni/thesis/gastric_ints.csv')
gastric_metadata <- nmr_csv |> select(patient, time, PPeal, diabetes) |> 
  mutate(sample_name = paste(patient, time, PPeal, diabetes, sep = '_'))
gastric_df <- nmr_csv |> mutate(sample_name = paste(patient, time, PPeal, diabetes, sep = '_'), .keep = 'unused') |> relocate(sample_name)

gastric_df_t <- gastric_df |> column_to_rownames("sample_name") |> t() |> as.data.frame()
gastric_df_t <- gastric_df_t[nrow(gastric_df_t):1, ]


gastric_norm <- normaliseSpectra(gastric_df_t, type.norm = "mean")
gastric_align <- alignSpectra(gastric_norm)
gastric_obj <- createSpectra(gastric_align)
```

run ASICS
```{r}
#gastric_asics <- ASICS(gastric_obj, ncores = 8)
gastric_asics <- ASICS(gastric_obj)
```

save the computed object because it takes hours (if you only use 1 cor, that is)
```{r}
#saveRDS(gastric_asics, file = '/Users/b246357/Documents/uni/thesis/gastric_asics.rds')
#gastric_asics <- readRDS('/Users/b246357/Documents/uni/thesis/gastric_asics.rds')
```


```{r}
plot(gastric_asics, idx = 1, add.metab = "Lactate")
```


gastric pca
```{r}
gastric_pca <- gastric_asics@quantification %>% t() %>% prcomp(scale. = T)


gastric_annotation <- gastric_metadata |> column_to_rownames("sample_name") |> mutate(gastric_annotation = paste(time, PPeal, diabetes, sep = '_'))
gastric_pca_df <- gastric_pca$x %>% as.data.frame() %>% bind_cols(gastric_annotation)
gastric_pca_df %>% ggplot(aes(x=PC1, y=PC2, col=diabetes))+geom_point(size=1)

```



```{r}
gastric_mat <- as.matrix(gastric_asics@quantification)
gastric_var <- apply(gastric_mat, 1, var, na.rm = TRUE)

# Keep top 30 most variable gastrics
top <- names(sort(gastric_var, decreasing = TRUE))[1:30]

mat_top <- gastric_mat[top, , drop = FALSE]


pheatmap::pheatmap(
  #gastric_mat,
  mat_top,
  annotation_col = select(gastric_annotation, -c(gastric_annotation, patient)),
  cluster_cols = TRUE,
  cluster_rows = TRUE,
  border_color = NA,
  fontsize_row = 7
)
```

