---
title: "metabodeconplus"
output:
  pdf_document: default
  html_document: default
date: "2026-09-16"
---

```{r setup, include=FALSE}
library(tidyverse)
library(magrittr)
library(metabodeconplus)
```

## simulated datafrom metabodecon packages


```{r cars}
sim_dir <- metabodeconplus::metabodeconplus_file("bruker/sim")
sim <- metabodeconplus::read_spectra(sim_dir)
deconvoluted_spectra_sim <- metabodeconplus::deconvolute(
    sim,                 # The object containing spectra
    sfr = c(3.35, 3.55), # Borders of signal free region (SFR) in ppm
    smit = 2, smws = 5,  # Smoothing parameters
    verbose = FALSE      # Disable verbose output
)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
# Visualize the first spectrum.
metabodeconplus::plot_spectrum(deconvoluted_spectra_sim[[1]])

# Visualize the second spectrum, this time without the legend.
metabodeconplus::plot_spectrum(deconvoluted_spectra_sim[[1]], lgd = FALSE)

# Visualize all spectra and save them to a pdf file
pdfpath <- tempfile(fileext = ".pdf")
pdf(pdfpath)
for (x in deconvoluted_spectra_sim) {
    metabodeconplus::plot_spectrum(x, main = x$filename)
}
dev.off()
cat("Plots saved to", pdfpath, "\n")
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
```{r}
# Plot spectra before alignment. Only show spectra 1-8 for clarity.
metabodeconplus::plot_spectra(deconvoluted_spectra_sim[1:8], lgd = FALSE)

# Align spectra and plot again.
aligned_spectra_sim <- metabodeconplus::align(deconvoluted_spectra_sim)
metabodeconplus::plot_spectra(aligned_spectra_sim[1:8])
```

## load own spectra
Note that the expected folder structure is a bit different for this package, but I was lazy so didn't wanna change my own folder structure, this is why they all end up having the same name.
```{r}
spectral_dir <- "/Users/b246357/Documents/uni/thesis/media_extraction_comparison/Lars260911"
spectra <- 1:9 %>% lapply(function(x){read_spectrum(spectral_dir, 'bruker', x, 1)}) %>% as_spectra()
deconvoluted_spectra <- metabodeconplus::deconvolute(
    spectra,                 # The object containing spectra
    sfr = c(0, 7.5), # Borders of signal free region (SFR) in ppm
    smit = 2, smws = 5,  # Smoothing parameters
    verbose = FALSE      # Disable verbose output
)
```


```{r}
# Visualize the first spectrum.
metabodeconplus::plot_spectrum(deconvoluted_spectra[[1]], foc_rgn = c(4.5,5.5))

# Visualize the second spectrum, this time without the legend.
#metabodeconplus::plot_spectrum(deconvoluted_spectra[[1]], foc_rgn = c(-0.5,7.5))
metabodeconplus::plot_spectra(deconvoluted_spectra, foc_rgn = c(-0.5,7.5))
# Visualize all spectra and save them to a pdf file
pdfpath <- tempfile(fileext = ".pdf")
pdf(pdfpath)
for (x in deconvoluted_spectra) {
    metabodeconplus::plot_spectrum(x, main = x$filename)
}
dev.off()
cat("Plots saved to", pdfpath, "\n")
```

check nr 7 coz' it's scuffed
```{r}
metabodeconplus::plot_spectrum(deconvoluted_spectra[[7]], foc_rgn = c(3,4))
test_spectrum <- spectra[[7]]
test_deconv <- deconvolute(test_spectrum)
```



Align spectra
```{r}
# Plot spectra before alignment. Only show spectra 1-8 for clarity.
metabodeconplus::plot_spectra(deconvoluted_spectra[1:9], foc_rgn = c(-0.2,7.5))

# Align spectra and plot again.
harmonized_spectra <- metabodeconplus::harmonize_grid(deconvoluted_spectra, target = "median", pad = 0)
aligned_spectra <- metabodeconplus::align(harmonized_spectra[-7]) # issues with spectrum 7 because of low intensity
#aligned_spectra <- metabodeconplus::align(deconvoluted_spectra[-7])
metabodeconplus::plot_spectra(aligned_spectra[1:8], foc_rgn = c(-0.2,7.5))
```

```{r}
binned <- metabodeconplus::bin(aligned_spectra)
pca_binned <- prcomp(binned)

pca_binned_df <- data.frame(pc1=pca_binned$x[1, ], pc2 = pca_binned$x[2,])
pca_binned_df %<>% mutate(extraction_method = c('E', 'K', 'E', 'E', 'K', 'K', 'E','E'),
                          media_type = c('unknown', '+cell', '+cell', '-cell', '-cell', '-cell', '-cell', '+cell'))
pca_binned_df %>% ggplot(aes(x = pc1, y= pc2, col = extraction_method, shape = media_type )) + geom_point()
```


signal integral matrix
```{r}
signalintegral <- metabodeconplus::si_mat(aligned_spectra)
pca_signalintegral <- prcomp(signalintegral)

pca_signalintegral_df <- data.frame(pc1=pca_signalintegral$x[1, ], pc2 = pca_signalintegral$x[2,])
pca_signalintegral_df %<>% mutate(extraction_method = c('E', 'K', 'E', 'E', 'K', 'K', 'E','E'),
                          media_type = c('unknown', '+cell', '+cell', '-cell', '-cell', '-cell', '-cell', '+cell'))
pca_signalintegral_df %>% ggplot(aes(x = pc1, y= pc2, col = extraction_method, shape = media_type )) + geom_point()
```

peak mat

```{r}
peak_mat <- metabodeconplus::si_mat(aligned_spectra)
pca_peak_mat <- prcomp(peak_mat)

pca_peak_mat_df <- data.frame(pc1=pca_peak_mat$x[1, ], pc2 = pca_peak_mat$x[2,])
pca_peak_mat_df %<>% mutate(extraction_method = c('E', 'K', 'E', 'E', 'K', 'K', 'E','E'),
                          media_type = c('unknown', '+cell', '+cell', '-cell', '-cell', '-cell', '-cell', '+cell'))
pca_peak_mat_df %>% ggplot(aes(x = pc1, y= pc2, col = extraction_method, shape = media_type )) + geom_point()
```

