---
title: "statistics"
output:
  pdf_document: default
  html_document: default
date: "2026-09-16"
---

```{r setup, include=FALSE}
library(tidyverse)
library(magrittr)
library(ASICS)
library(pheatmap)
```

## heatmap visualization


```{r cars}
library(dplyr)
library(tidyr)
library(ggplot2)
library(pheatmap)

set.seed(42)

# Experimental design
subjects <- paste0("S", sprintf("%02d", 1:20))

metadata <- data.frame(
  subject = subjects,
  condition = rep(c("Control", "Treatment"), each = 10)
)

metabolites <- c(
  "Glucose", "Lactate", "Alanine", "Acetate",
  "Glutamate", "Glutamine", "Citrate", "Pyruvate",
  "Valine", "Leucine", "Isoleucine", "Choline"
)

times <- c("T0", "Week1", "Month1")

# Generate BEFORE measurements
toy <- expand_grid(
  subject = subjects,
  time = times,
  metabolite = metabolites
) |>
  left_join(metadata, by = "subject") |>
  mutate(
    before = rlnorm(
      n(),
      meanlog = 2,
      sdlog = 0.3
    )
  )

toy <- toy |>
  mutate(
    true_log2fc = case_when(

      # Glucose decreases after meal
      metabolite == "Glucose" ~ -0.8,

      # Lactate increases
      metabolite == "Lactate" ~ 0.7,

      # Treatment progressively increases acetate response
      metabolite == "Acetate" &
        condition == "Treatment" &
        time == "T0" ~ 0.2,

      metabolite == "Acetate" &
        condition == "Treatment" &
        time == "Week1" ~ 0.8,

      metabolite == "Acetate" &
        condition == "Treatment" &
        time == "Month1" ~ 1.4,

      # Treatment changes glutamate response
      metabolite == "Glutamate" &
        condition == "Treatment" ~ 0.8,

      TRUE ~ 0
    ),

    # Add biological noise to meal response
    observed_log2fc =
      true_log2fc + rnorm(n(), 0, 0.25),

    # Generate AFTER measurement
    after =
      before * 2^observed_log2fc
  )

toy <- toy |>
  mutate(
    log2FC = log2(after / before)
  )
```

generate a toy FC matrix


```{r pressure}
fc_matrix <- toy |>
  select(
    subject,
    condition,
    time,
    metabolite,
    log2FC
  ) |>
  unite(
    "sample",
    subject,
    time,
    remove = FALSE
  ) |>
  select(sample, metabolite, log2FC) |>
  pivot_wider(
    names_from = metabolite,
    values_from = log2FC
  ) |>
  tibble::column_to_rownames("sample") |>
  as.matrix()

```


## not very useful heatmap
```{r}
annotation <- toy |>
  distinct(subject, condition, time) |>
  unite("sample", subject, time, remove = FALSE) |>
  select(sample, condition, time) |>
  tibble::column_to_rownames("sample") |>
  as.data.frame()

annotation$condition <- factor(annotation$condition)
annotation$time <- factor(
  annotation$time,
  levels = c("T0", "Week1", "Month1")
)

pheatmap(
  t(fc_matrix),
  annotation_col = annotation,
  scale = "none",
  border_color = NA,
  clustering_distance_cols = "euclidean",
  clustering_distance_rows = "euclidean"
)
```

3 heatmaps instead?

```{r}
library(dplyr)
library(tidyr)
library(pheatmap)
library(gridExtra)

make_fc_heatmap <- function(df, timepoint, annotation_cols = NULL) {
  mat <- df |>
    filter(time == timepoint) |>
    select(subject, metabolite, log2FC) |>
    pivot_wider(names_from = metabolite, values_from = log2FC) |>
    tibble::column_to_rownames("subject") |>
    as.matrix()

  ann <- df |>
    filter(time == timepoint) |>
    distinct(subject, condition) |>
    tibble::column_to_rownames("subject")

  pheatmap(
    t(mat),
    annotation_col = ann,
    scale = "none",
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    border_color = NA,
    silent = TRUE
  )$gtable
}

h0 <- make_fc_heatmap(toy, "T0")
h1 <- make_fc_heatmap(toy, "Week1")
h2 <- make_fc_heatmap(toy, "Month1")

gridExtra::grid.arrange(h0, h1, h2, ncol = 3)
```




PCA
```{r}
pca <- prcomp(fc_matrix, center = TRUE, scale. = TRUE)

scores <- as.data.frame(pca$x[, 1:2]) |>
  tibble::rownames_to_column("sample") |>
  left_join(annotation |> tibble::rownames_to_column("sample"),
            by = "sample") |> mutate(condition_time = paste(condition, time, sep = '_'))

ggplot(scores, aes(PC1, PC2, color = condition, shape = time)) +
  geom_point(size = 3) +
  theme_classic()

```

heatmap with same ordering

```{r}
library(dplyr)
library(tidyr)
library(pheatmap)
library(gridExtra)

plot_fc_heatmaps <- function(df, times = c("T0", "Week1", "Month1"),
                             top_n = 30) {

  # Select most variable metabolites across all time points
  top_metabolites <- df |>
    group_by(metabolite) |>
    summarise(
      variance = var(log2FC, na.rm = TRUE),
      .groups = "drop"
    ) |>
    arrange(desc(variance)) |>
    slice_head(n = top_n) |>
    pull(metabolite)

  df <- df |> filter(metabolite %in% top_metabolites)

  # Common scale
  max_abs <- max(abs(df$log2FC), na.rm = TRUE)
  breaks <- seq(-max_abs, max_abs, length.out = 101)

  cols <- colorRampPalette(
    c("blue", "white", "red")
  )(100)

  # Common row order
  row_order <- df |>
    group_by(metabolite) |>
    summarise(mean_fc = mean(log2FC, na.rm = TRUE), .groups = "drop") |>
    arrange(mean_fc) |>
    pull(metabolite)

  make_plot <- function(tp) {

    mat <- df |>
      filter(time == tp) |>
      select(subject, metabolite, log2FC) |>
      pivot_wider(names_from = metabolite, values_from = log2FC) |>
      tibble::column_to_rownames("subject") |>
      as.matrix()

    mat <- t(mat)
    mat <- mat[row_order[row_order %in% rownames(mat)], , drop = FALSE]

    ann <- df |>
      filter(time == tp) |>
      distinct(subject, condition) |>
      tibble::column_to_rownames("subject")

    pheatmap(
      mat,
      annotation_col = ann,
      cluster_rows = FALSE,
      cluster_cols = TRUE,
      scale = "none",
      breaks = breaks,
      color = cols,
      border_color = NA,
      main = tp,
      silent = TRUE
    )$gtable
  }

  gridExtra::grid.arrange(
    grobs = lapply(times, make_plot),
    ncol = length(times)
  )
}

plot_fc_heatmaps(toy, top_n = 30)
```

volcano plot idea, note that the p.value in this case might be different from the emmeans one below
```{r}
library(dplyr)
library(ggplot2)
library(patchwork)

make_fc_volcano <- function(df, timepoint,
                            label_n = 10) {

  stats <- df |>
    filter(time == timepoint) |>
    group_by(metabolite) |>
    summarise(
      control_fc = mean(log2FC[condition == "Control"], na.rm = TRUE),
      treatment_fc = mean(log2FC[condition == "Treatment"], na.rm = TRUE),
      log2_fc_fc = treatment_fc - control_fc,
      p_value = t.test(
        log2FC[condition == "Control"],
        log2FC[condition == "Treatment"]
      )$p.value,
      .groups = "drop"
    ) |>
    mutate(
      FDR = p.adjust(p_value, method = "BH"),
      neg_log10_p = -log10(p_value)
    )

  labels <- stats |>
    arrange(FDR) |>
    slice_head(n = label_n)

  ggplot(stats, aes(log2_fc_fc, neg_log10_p)) +
    geom_point() +
    geom_vline(xintercept = 0, linetype = "dashed") +
    geom_text(
      data = labels,
      aes(label = metabolite),
      vjust = -0.7,
      size = 3
    ) +
    labs(
      title = timepoint,
      x = "log2(fold-change of fold-change)",
      y = "-log10(p-value)"
    ) +
    theme_classic()
}

p0 <- make_fc_volcano(toy, "T0")
p1 <- make_fc_volcano(toy, "Week1")
p2 <- make_fc_volcano(toy, "Month1")

p0 + p1 + p2
```

individual metabolite trajectories, note that these need to be fixed so that the correct chronology is shown (t0, w1, month0), chatGPT isn't aware of time
```{r}
toy |>
  filter(metabolite %in% c(
    "Lactate", "Glucose", "Alanine", "Acetate"
  )) |>
  ggplot(
    aes(
      x = time,
      y = log2FC,
      group = subject,
      color = condition
    )
  ) +
  geom_line(alpha = 0.2) +
  stat_summary(
    aes(group = condition),
    fun = mean,
    geom = "line",
    linewidth = 1
  ) +
  stat_summary(
    aes(group = condition),
    fun = mean,
    geom = "point",
    size = 2
  ) +
  facet_wrap(~ metabolite, scales = "free_y") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic()
```

## lme
```{r}
library(lme4)
library(lmerTest)
fit <- lmer(
  log2FC ~ condition * time + (1 | subject),
  data = toy
)
```


```{r}
library(dplyr)
library(lme4)
library(lmerTest)
library(broom.mixed)

results <- toy |>
  group_by(metabolite) |>
  group_modify(\(d, ...) {
    fit <- lmer(
      log2FC ~ condition * time + (1 | subject),
      data = d
    )

    broom.mixed::tidy(fit, effects = "fixed")
  }) |>
  ungroup()


```


  
```{r}
library(lme4)
library(lmerTest)
library(emmeans)
library(dplyr)

results <- toy |>
  group_by(metabolite) |>
  group_modify(\(d, ...) {

    fit <- lmer(
      log2FC ~ condition * time + (1 | subject),
      data = d
    )

    emm <- emmeans(
      fit,
      ~ condition | time
    )

    as.data.frame(
      pairs(emm)
    )
  }) |>
  ungroup()

print(results)
```

sanity check
```{r}
d <- toy |>
  filter(metabolite == "Acetate")

fit <- lmer(
  log2FC ~ condition * time + (1 | subject),
  data = d
)

emm <- emmeans(
  fit,
  ~ condition | time
)

# All condition comparisons
pairs(emm)

# Month1 only
pairs(emm) |>
  as.data.frame() |>
  filter(time == "Month1")

d |>
  filter(time == "Month1") |>
  group_by(condition) |>
  summarise(
    mean_log2FC = mean(log2FC),
    sd = sd(log2FC),
    n = n()
  )
```


